---
title: "Comparing the Presidential Election with the 2024 Paid Sick Leave Referendum"
output: html_document
date: "2026-01-08"
author: "John Verghese"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(sf)
library(knitr)
library(kableExtra)
library(ggrepel)
library(gt)
library(plotly)

pres <- read_csv("ne_pres_cleaned.csv")
abor <- read_csv("ne_439_cleaned.csv")
paid <- read_csv("ne_436_cleaned.csv")

cens <- read_sf("ne_census_wide.gpkg")
cens_long <- read_sf("ne_census_long.gpkg")

elec <- pres |> 
  inner_join(paid, by="county", suffix = c("_pres", "_paid")) |>
  mutate(split_margin = votes_for - votes_kamala)

elec_cens_long <- cens_long |> inner_join(elec, by="county")
  
elec_cens_geo <- cens |> inner_join(elec, by="county")
elec_cens <- st_drop_geometry(elec_cens_geo) |>
  select(-c("GEOID","NAME", "acs_male", "acs_female")) |>
  relocate(county, acs_total_pop, votes_totals_pres, votes_totals_paid, acs_median_income) |>
  mutate(across(-county, ~ if_else(.x < 1, .x * 100, .x)))

```

  Paid Sick Leave for Nebraskans and a table of progressive-allied groups, including the AFL-CIO, Planned Parenthood, and the ACLU, supported a referendum mandating paid sick leave for all working Nebraskans. Initiative 436 went on the ballot in the 2024 general election, sharing the ballot with the high-salience presidential election, downballot races, and a number of other ballot measures, including Initiative 439 to codify a woman's right to choose.
  Initiative 436 proved to be the most popular item on the ballot. Nearly 75% of Nebraskans supported the measure^[Thanks to the NE Secretary of State and the US Census Bureau for all data used to generate this report], with all but four counties (which are among the least populous in the state) returning a majority for the measure:
``` {r broad_support_table, echo = FALSE}
elec_cens |> 
  filter(votes_for < 50) |>
  select(county, votes_for, acs_total_pop) |>
  arrange(votes_for) |>
  add_row(elec_cens |> 
            filter(votes_for > 50) |> 
            summarize(county = "All Other Counties", 
                      votes_for = weighted.mean(x = votes_for, w = votes_totals_paid),
                      acs_total_pop = sum(acs_total_pop)
                      )
          ) |>
  gt( rowname_col = "county") |>
  fmt_percent(columns = "votes_for", 
                 scale_values = FALSE,
                 decimals = 1) |>
  fmt_number(columns = "acs_total_pop", decimals = 0) |>
  cols_label(votes_for ~ "Init 436 Support", acs_total_pop ~ "Population") 
    

```
This overwhelming support demonstrates the popularity of progressive economic measures to support working Nebraskans. In every county in the state, Initiative 436 significantly overperformed the top of the Democratic ticket:

``` {r overperformance, fig.height = 15, echo = FALSE}
elec_cens |> 
  ggplot() + 
  geom_point(aes( x = votes_kamala, y = fct_reorder(county, desc(votes_kamala)))) + 
  geom_segment(aes( 
    x = votes_kamala, 
    xend = votes_for, 
    y = fct_reorder(county, desc(votes_for))), 
    arrow = arrow(length = unit(0.05, "inches"))
  ) +
  labs(title = "Paid Leave Beats Top-Ticket Dems by ~40 points") +
  xlab("Increase in support between Harris/Walz and Initiative 436") +
  ylab("")
```



  Omaha and Lincoln metropolitan counties of Douglas/Sarpy and Lancaster counties, in addition to the majority-Indigenous Thurston County returned the strongest support for the Democratic ticket and the strongest majorities in support of Initiative 436:

```{r corr_plot, echo = FALSE}
ggplot(elec_cens, aes(x = votes_kamala, y = votes_for)) +
  geom_point(aes(color = county, size = acs_total_pop), show.legend = FALSE, alpha = 0.5) +
  geom_text_repel(
    data = elec_cens |> 
      filter(
        county %in% c(
        "Thurston", 
        "Sarpy", 
        "Douglas", 
        "Lancaster", 
        "Keya Paha")
        ), 
    aes(
      x = votes_kamala, 
      y = votes_for, 
      label=county
      ), 
    position = "jitter",
    arrow = arrow(length = unit(0.25, "lines"))
    ) +
  labs(
    title = "Top-Ticket Dem v Initiative 436 Support",
    x = "% Harris/Walz Support",
    y = "% Yes on Initiative 436"
  ) +
  xlim(0,100) + 
  ylim(0,100)
```

There was a strong racial and age correlation with support for the ballot measure, with young, black, and Hispanic counties returning significant support:

```{r facet_plots, warning = FALSE, message = FALSE, echo=FALSE}
elec_cens_long |>
  filter(!acs_var %in% c("acs_male", "acs_female")) |>
  mutate(acs_var = str_to_title(str_replace(str_remove(acs_var, "acs_"), "_", " "))) |>
  ggplot(aes(x = votes_for, y = acs_est)) + 
    geom_point(aes(color = county, size = votes_totals_pres), show.legend = FALSE, alpha = 0.5) +
    xlim(0,1) +
    geom_smooth(se = FALSE) +
    facet_wrap(~acs_var, scales = "free_y") +
    labs(x = "Initiative 436 Support", y = "")
```

In the five counties with more than a quarter of the population identifying as Latino - Colfax, Dakota, Dawson, Hall, and Saline - paid sick leave beat the top of the ticket by ~45 points:

``` {r hispanic_counties, echo=FALSE}
elec_cens |> 
  arrange(desc(acs_hispanic)) |> 
  head(5) |> 
  select(county, acs_total_pop, votes_kamala, votes_for, acs_hispanic) |> 
  relocate(county, acs_hispanic) |> 
  mutate(split_margin = votes_for - votes_kamala) |>
  gt( rowname_col = "county") |>
  fmt_percent(columns = c("votes_for", "votes_kamala", "split_margin", "acs_hispanic"),  
                 scale_values = FALSE,
                 decimals = 1) |>
  fmt_number(columns = "acs_total_pop", decimals = 0) |>
  cols_label(votes_for ~ "Init 436 Support", 
             acs_total_pop ~ "Population",
             acs_hispanic ~ "% Latino",
             votes_kamala ~ "Harris Support",
             split_margin ~ "Difference") 
```

The takeaway: voters in Nebraska across geographies have indicated a willingness to support progressive economic policies. Progressives in Nebraska would do well to strongly associate themselves with the success of Initiative 436, and continue to focus on economically progressive policies.


``` {r votes_on_table, echo = FALSE}
elec_cens |> 
  mutate(county_type = case_when
         (county %in% c("Douglas", "Lancaster", "Sarpy") ~ "Metro", 
          county == "Thurston" ~ "Thurston", 
           county %in% c("Colfax", "Dakota", "Dawson", "Hall", "Saline") ~ "Latino > 25%", 
           .default = "Rest of State")) |> 
  group_by(county_type) |> 
  summarize(total_votes = sum(votes_totals_pres), 
            votes_for_t = sum((votes_for/100) * votes_totals_paid), 
            votes_kamala_t = sum((votes_kamala/100) * votes_totals_pres),
            votes_on_table = sum((votes_for/100) * votes_totals_paid) - sum((votes_kamala/100) * votes_totals_pres)) |>
  gt(rowname_col = "county_type") |> 
  fmt_number(columns = c("total_votes", "votes_for_t", "votes_kamala_t", "votes_on_table"), decimals = 0) |>
  cols_label(total_votes = "Total Voters",
             votes_for_t = "Initiative 436 Supporters",
             votes_kamala_t = "Top-ticket Dem Supporters",
             votes_on_table = "Swingable Voters")
  

```



See the geographic distribution of increased support for paid sick leave vs the Harris/Walz ticket below:

```{r map, fig.width = 15, out.width = "100%", warning = FALSE,  echo = FALSE}

ggplot(elec_cens_geo) +
  geom_sf(aes(fill = split_margin), color = "white", size = 0.3) +
  scale_fill_gradient2(
    low = "white",
    mid = "#ffffff", 
    high = "#2b83ba",
    midpoint = 0.3,
    name = "Overperformance of Paid Leave Policy vs Dem Ticket",
    labels = function(x) paste0(ifelse(x > 0, "+", ""), x * 100, "%")
  ) +
  labs(
    title = "Geographic Distribution of Split-Ticket Voting",
    subtitle = "Blue = Increasing support for Initiative 436 relative to Harris/Walz",
    caption = "Nebraska 2024"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "right"
  ) + 
  geom_sf_text(aes(label = county), position = position_dodge2())
```

Additional analysis is available upon request, including comparison of Initiative 439, on a woman's right to choose, with the paid sick leave and presidential elections.

_AI Disclaimer: I wrote all of the code used to generate this document, with the exception of choosing which census variables to pull into the analysis._

